---
title: "ETM58D Business Analytics Final"
author: "Mehmet Arda Keskin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    number_sections: true
---

# Setup and Loading Libraries

```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE,fig.align = "center")
options(scipen = 999, knitr.kable.NA = "")

# Installing libraries (if do not have)
##install.packages("tidyverse")
##install.packages("lubridate")
##install.packages("readxl")
##install.packages("skimr")
##install.packages("magrittr")
##install.packages("tidyquant")
##install.packages("tsibble")
##install.packages("feasts")
##install.packages("ggcorrplot")
##install.packages("glmnet")
##install.packages("caret")
##install.packages("rattle")

# Importing libraries
library(tidyverse)
library(lubridate)
library(readxl)
library(skimr)
library(magrittr)
library(tidyquant)
library(tsibble)
library(feasts)
library(ggcorrplot)
library(glmnet)
library(caret)
library(rattle)

set.seed(1234)
```

# Import Data

Electricity price, consumption and production data is imported from csv file.

I have chosen the dollars for the prediction currency due to the fact that the US dollar is one of the most widely used, generally accepted and stable currencies in the world. Other currencies especially TRY may fluctuate greatly over the years due to irrelevant factors of this analysis topic such as inflation, interest, etc.

```{r}
df <- read_csv("mcp_with_variables.csv") %>% select(-mcp_try, -mcp_euro)
head(df)
```

# Descriptive Analysis

Summary statistics of variables are shown.

We can say that **mcp_dollars** have some extreme values (p100 = 542). Also, it can be stated that **load_plan** and **total_prod** have similar distribution from their histogram.

```{r}
skim(df)
```

Histogram and Box plot of **mcp_dollars** variable are shown.

```{r}
ggplot(df, aes(x=mcp_dollars)) + geom_histogram()
boxplot(df$mcp_dollars)
```

Histogram **load_plan** and **total_prod** variable are shown. We can say that variables are distributed similarly and have characteristics of normal distribution around the mean value of 30000.

```{r}
ggplot(df, aes(x=load_plan)) + geom_histogram()
ggplot(df, aes(x=total_prod)) + geom_histogram()
```

Box plots of load_plan total_prod variable

```{r}
boxplot(df$load_plan)
boxplot(df$total_prod)
```

Seasonality plots of **load_plan** variable for several hours are shown. 0, 4, 8, 12, 16, 20 values of hours are tried to be plotted. It can be said that there is generally some trend pattern yearly. However, it is seen that it shows some other seasonal pattern for a shorter period of time like weekly or monthly.


```{r}
as_tsibble(df %>% filter(hour == 0)) %>% gg_season(load_plan) + theme_tq()
as_tsibble(df %>% filter(hour == 4)) %>% gg_season(load_plan) + theme_tq()
as_tsibble(df %>% filter(hour == 8)) %>% gg_season(load_plan) + theme_tq()
as_tsibble(df %>% filter(hour == 12)) %>% gg_season(load_plan) + theme_tq()
as_tsibble(df %>% filter(hour == 16)) %>% gg_season(load_plan) + theme_tq()
as_tsibble(df %>% filter(hour == 20)) %>% gg_season(load_plan) + theme_tq()
```

Four weeks of MCP Price line graph for every hour is shown to check if there is any seasonality in the variable. There are some weekly seasonality for some hours but there is no general seasonality for whole day.

```{r}
df %>%
  filter(date >= as_date("2019-01-07") & date <= as_date("2019-02-04")) %>% 
  ggplot(aes(x = date, y = mcp_dollars)) + 
  geom_line(aes(color = as.factor(hour)), size = 0.9, show.legend = T) + 
  labs(title="", y="MCP Price (USD)", x="Date") + 
  facet_wrap(~hour, ncol=4) + 
  theme_tq()
```

Hourly MCP Price line graph for every day of a week is shown to check if there is any seasonality in the variable. There are some similar patterns for some days of weeks; however, there are not exact pattern for every day.

```{r}
df %>%
  filter(date >= as_date("2019-01-07") & date <= as_date("2019-01-13")) %>% 
  ggplot(aes(x = hour, y = mcp_dollars)) + 
  geom_line(aes(color = as.factor(date)), size = 0.9) + 
  labs(title="", y="MCP Price (USD)", x="Hour") + 
  theme_tq()
```

Four weeks Electricity Usage (load_plan) line graph for every hour is shown to check if there is any seasonality in the variable. We can absolutely say that there is weekly seasonality pattern in this variable. Every same day of week shows similar pattern and especially Sundays, there are a sharp decrease in electricity usage.

```{r}
df %>%
  filter(date >= as_date("2019-01-07") & date <= as_date("2019-02-04")) %>% 
  ggplot(aes(x = date, y = load_plan)) + 
  geom_line(aes(color = as.factor(hour)), size = 0.9, show.legend = T) + 
  labs(title="", y="Electricity Usage", x="Date") + 
#  facet_wrap(~hour, ncol=4) + 
  theme_tq()
```

Hourly Electricity Usage (load_plan) line graph for every day of a week is shown to check if there is any seasonality. We can also absolutely say that there is hourly seasonality pattern in this variable. Every same hour of a day shows similar pattern. Electricity usage of night hours are low as it is expected.

```{r}
df %>%
  filter(date >= as_date("2019-01-07") & date <= as_date("2019-01-13")) %>% 
  ggplot(aes(x = hour, y = load_plan)) + 
  geom_line(aes(color = as.factor(date)), size = 0.9) + 
  labs(title="", y="Electricity Usage", x="Hour") + 
  theme_tq()
```

Four weeks Total Electricity Production (total_prod) line graph for every hour is shown to check if there is any seasonality. It shows similar characteristics with electricity usage (load_plan) variable.

```{r}
df %>%
  filter(date >= as_date("2019-01-07") & date <= as_date("2019-02-04")) %>% 
  ggplot(aes(x = date, y = total_prod)) + 
  geom_line(aes(color = as.factor(hour)), size = 0.9, show.legend = T) + 
  labs(title="", y="Total Electricity Production", x="Date") + 
#  facet_wrap(~hour, ncol=4) + 
  theme_tq()
```

Hourly Total Electricity Production (total_prod) line graph for every day of a week is shown to check if there is any seasonality. It shows similar characteristics with electricity usage (load_plan) variable.

```{r}
df %>%
  filter(date >= as_date("2019-01-07") & date <= as_date("2019-01-13")) %>% 
  ggplot(aes(x = hour, y = total_prod)) + 
  geom_line(aes(color = as.factor(date)), size = 0.9) + 
  labs(title="", y="Total Electricity Production", x="Hour") + 
  theme_tq()
```

# Data Manipulation

New variables like lagged variables or ratio type variables are created to normalize the seasonality/trend effects. **renew_energy_ratio** variable shows ratio of the production fulfilled by renewable energy sources. **cons_prod_ratio** variable shows how much total production meets demand. Finally, weekly (lag168) and monthly (lag672) lagged variables created from both of these new variables and also from the target variable.

```{r}
df2 <- df %>%
  mutate(renew_energy_ratio = (wind + geothermal + dam + biomass + river) / total_prod,
         cons_prod_ratio = load_plan / total_prod,
         mcp_dollars_lag168 = lag(mcp_dollars, 168),
         renew_energy_ratio_lag168 = lag(renew_energy_ratio, 168),
         cons_prod_ratio_lag168 = lag(cons_prod_ratio, 168),
         mcp_dollars_lag672 = lag(mcp_dollars, 672),
         renew_energy_ratio_lag672 = lag(renew_energy_ratio, 672),
         cons_prod_ratio_lag672 = lag(cons_prod_ratio, 672)) %>% 
  select(date, hour, mcp_dollars, mcp_dollars_lag168, renew_energy_ratio_lag168, cons_prod_ratio_lag168, 
         mcp_dollars_lag672, renew_energy_ratio_lag672, cons_prod_ratio_lag672) %>% 
  filter(date >= as_date("2015-06-29"))
```

Variable distribution, scatter and histogram plots are drawn for one week with selected variables

```{r}
GGally::ggpairs(df2 %>% filter(date >= as_date("2019-01-07") & date <= as_date("2019-01-13")) %>% select(-date, -hour)) + theme_tq()
```

Train and test data sets are created

```{r}
df2_train <- df2 %>% filter(date < as_date("2021-05-22"))
df2_test <- df2 %>% filter(date >= as_date("2021-05-22"))
```

# Modeling

```{r}
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 5)
```

Linear Regression model with repeated cross validation (fitControl) is created by using caret package ("lm" method). Summary results of the model is shown. R-squared value is not high enough. The explanatory power of the model seems low at first glance.

```{r}
lin_reg <- train(mcp_dollars ~ ., 
                 data = df2_train %>% select(-date, -hour),
                 method = "lm",
                 trControl = fitControl,
                 tuneLength = 5)
summary(lin_reg)
```

Decision Tree model with repeated cross validation (fitControl) is created by using caret package ("rpart" method). RMSE vs Complexity Parameter plot is shown and the model with lowest CP and RMSE is chosen as a final model among the decision trees. Also, the tree plot is shown and it can be said that only lag variables of mcp_dollars is used at the best decision tree alternative.

```{r}
dec_tree <- train(mcp_dollars ~ ., 
                  data = df2_train %>% select(-date, -hour),
                  method = "rpart",
                  trControl = fitControl,
                  tuneLength = 5)
trellis.par.set(caretTheme())
plot(dec_tree) 
fancyRpartPlot(dec_tree$finalModel)
```

Random Forest model with repeated cross validation (fitControl) is created by using caret package ("ranger" method). Impurity measure is used as an importance criteria and number of trees parameter is selected as 50 because of performance issues. Final model of random forest have value of 2 as mtry parameter and splitrule as variance. RMSE value of best random forest alternative is 9.10 and R-squared value of best random forest alternative is around 60% which is better than other model methods (linear regression and decision tree).

```{r}
random_forest <- train(mcp_dollars ~ ., 
                      data = df2_train %>% select(-date, -hour),
                      method = "ranger",
                      trControl = fitControl,
                      num.trees = 50,
                      importance = "impurity")
plot(random_forest)
random_forest$results
random_forest$finalModel
```

Candidate models are compared below. It is obviously seen that Random Forest model have better R-squared value and lower (so better) MAE and RMSE values. As a result of these analysis, Random Forest model is selected.

```{r}
results = resamples(list(Linear_Regression = lin_reg, Decision_Tree = dec_tree, Random_Forest = random_forest))
summary(results)
bwplot(results)
```

Prediction and residual analysis for train set of random forest model is shown. Actual vs Predicted plot shows that predicted values are generally close to the actual values. There are few observations which are too much deviated. Histogram of residuals shows normal distribution around mean value of 0 which is good for the model. Finally, predicted vs residuals plot shows that predicted values are distributed around 0 which is also good for the model even if there some few deviated values, they are not so important and not so much.

```{r}
rf_predicted <- predict(random_forest, data = df2_train)
plot(rf_predicted, df2_train$mcp_dollars, xlab = "Predicted", ylab = "Actual", main = "Actual vs Predicted Plot for Random Forest Model")
abline(a=0,b=1,col='red', lty = 2)

rf_residuals <- df2_train$mcp_dollars - rf_predicted
hist(rf_residuals, xlab = "Residuals", main = "Residuals Histogram of Random Forest Model")
plot(rf_predicted, rf_residuals, xlab = "Predicted", ylab = "Residuals", main = "Predicted vs Residuals Plot for Random Forest Model")
abline(h = 0, col = "red", lty = 2)
```

Variable Importance of random forest model is shown. **mcp_dollars_lag168** variable is the most important variable for the model followed by **mcp_dollars_lag672** and **renew_energy_ratio_lag168** variables.

```{r}
sort(random_forest$finalModel$variable.importance, decreasing = TRUE)
```

# Forecast for the Test Period

For the test period of data, target variable (mcp_dollars) are forecasted by using predict function. Prediction and residual analysis for test set of random forest model is shown. Actual vs Predicted plot shows that predicted values are generally close to the actual values. There are few observations which are too much deviated. Histogram of residuals shows normal distribution around mean value of 0 which is good for the model. Finally, predicted vs residuals plot shows that predicted values are distributed around 0 which is also good for the model even if there some few deviated values, they are not so important and not so much. 

```{r}
rf_predicted_test <- predict(random_forest$finalModel, data = df2_test)
plot(rf_predicted_test$predictions, df2_test$mcp_dollars, xlab = "Predicted", ylab = "Actual", main = "Actual vs Predicted Plot for Random Forest Model with Test Set")
abline(a=0,b=1,col='red', lty = 2)

rf_residuals_test <- df2_test$mcp_dollars - rf_predicted_test$predictions
hist(rf_residuals_test, xlab = "Residuals", main = "Residuals Histogram of Random Forest Model")
plot(rf_predicted_test$predictions, rf_residuals_test, xlab = "Predicted", ylab = "Residuals", main = "Predicted vs Residuals Plot for Random Forest Model with Test Set")
abline(h = 0, col = "red", lty = 2)
```

Lastly, RMSE values of test set predictions are shown. It is not a high value which is also good for the model.

```{r}
RMSE(df2_test$mcp_dollars, rf_predicted_test$predictions)
```

