---
title: "ETM58D Business Analytics Homework 2"
author: "Mehmet Arda Keskin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    number_sections: true
---

# Setup and Loading Libraries

```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE,fig.align = "center")
options(scipen = 999, knitr.kable.NA = "")

##install.packages("tidyverse")
##install.packages("lubridate")
##install.packages("readxl")
##install.packages("magrittr")
##install.packages("scatterplot3d")
##install.packages("ggfortify")

library(tidyverse)
library(lubridate)
library(readxl)
library(magrittr)
library(scatterplot3d)
library(ggfortify)
```

# TASK 1

Importing the csv file which show distance between the cities of Turkey.

```{r}
df_t1 <- read_csv("HW2_q1_data.csv") %>% as_tibble()
head(df_t1)
```

Replacing NAs with zeros (0)

```{r}
df_t1[is.na(df_t1)] <- 0
head(df_t1)
```

Applying multidimensional scaling (MDS)

```{r}
df_t1_fit <- cmdscale(df_t1, k = 2, eig = TRUE)
df_t1_fit
x_t1 <- df_t1_fit$points[, 1]
y_t1 <- df_t1_fit$points[, 2]
```

Representing cities on a 2D plot using the reduced dimensions

```{r}
plot(x_t1, y_t1, pch = 19)
text(x_t1, y_t1, pos = 4, cex = 0.6, labels = colnames(df_t1))
```

As an interesting result, plot shapes the map of Turkey with dots representing cities. Cities are almost on their exact locations.

# TASK 2

Importing the Netflix rating data and movie titles

```{r}
df_t2 <- read.table("HW2_q3_Netflix_data.dat", header=FALSE)
head(df_t2)
titles_t2 <- read_delim("HW2_q3_movie_titles.txt", delim = "/", col_names = FALSE)
head(titles_t2)
```

Checking missing values. It can be seen that there are not any missing values. So I assume that 0 ratings show the movie which are not rated by that user. I manipulate the data and change the **0** values to **NA**.

```{r}
df_t2[is.na(df_t2)]
df_t2[df_t2==0] <- NA
head(df_t2)
```

Imputing missing entries with the mean of the ratings for a particular movie

```{r}
df_t2 %<>% mutate_all(~ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))          
head(df_t2)
```

Creating euclidean distance matrix from the rating data by taking means of ratings for each movie

```{r}
df_t2_dist_matrix <- dist(colMeans(df_t2), method = "euclidean", upper = TRUE)
```

Applying multidimensional scaling (MDS)

```{r}
df_t2_fit <- cmdscale(df_t2_dist_matrix, k = 2, eig = TRUE)
df_t2_fit
x_t2 <- df_t2_fit$points[, 1]
y_t2 <- df_t2_fit$points[, 2]
```

Representing movies on a 2D plot using the reduced dimensions

```{r}
plot(x_t2, y_t2, pch = 16)
text(x_t2, y_t2, pos = 4, cex = 0.5, labels = titles_t2$X1)
```

As an interesting result, the movies which are in the IMDB top 250 list are generally located nearly and on the left side of the plot.


# TASK 3

Importing the dataset

```{r}
df_t3_x <- read.table("uWaveGestureLibrary_X_TRAIN", header=FALSE)
df_t3_y <- read.table("uWaveGestureLibrary_Y_TRAIN", header=FALSE)
df_t3_z <- read.table("uWaveGestureLibrary_Z_TRAIN", header=FALSE)
```

Data manipulation by converting data to long format (as an example first 6 rows of X axis data shown)

```{r}
df_t3_x_long <- df_t3_x %>% 
  as_tibble() %>% 
  mutate(time_series_id = row_number()) %>% 
  rename(class = V1) %>% 
  pivot_longer(cols = -c(time_series_id, class), names_to = "time_index", values_to = "X")

df_t3_y_long <- df_t3_y %>% 
  as_tibble() %>% 
  mutate(time_series_id = row_number()) %>% 
  rename(class = V1) %>% 
  pivot_longer(cols = -c(time_series_id, class), names_to = "time_index", values_to = "Y")

df_t3_z_long <- df_t3_z %>% 
  as_tibble() %>% 
  mutate(time_series_id = row_number()) %>% 
  rename(class = V1) %>% 
  pivot_longer(cols = -c(time_series_id, class), names_to = "time_index", values_to = "Z")

head(df_t3_x_long)
```

Binding columns to concatenate the time series of X, Y and Z axis. Summary of X, Y, Z columns are shown below.

```{r}
df_t3_long <- bind_cols(df_t3_x_long, Y = df_t3_y_long$Y, Z = df_t3_z_long$Z) %>% 
  mutate(class = as.factor(class))
head(df_t3_long)
summary(df_t3_long %>% select(X,Y,Z))
```

Selecting one instance from each class

```{r}
df_t3_class_1 <- df_t3_long %>% filter(time_series_id == 11 )
df_t3_class_2 <- df_t3_long %>% filter(time_series_id == 15 )
df_t3_class_3 <- df_t3_long %>% filter(time_series_id == 4 )
df_t3_class_4 <- df_t3_long %>% filter(time_series_id == 5 )
df_t3_class_5 <- df_t3_long %>% filter(time_series_id == 2 )
df_t3_class_6 <- df_t3_long %>% filter(time_series_id == 1 )
df_t3_class_7 <- df_t3_long %>% filter(time_series_id == 7 )
df_t3_class_8 <- df_t3_long %>% filter(time_series_id == 6 )
```

Visualizing the gestures from each class (Subtask a)

```{r}
scatterplot3d(x=df_t3_class_1$X, 
              y=df_t3_class_1$Y, 
              z=df_t3_class_1$Z, 
              main = "3D Scatter Plot - Class 1",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_2$X, 
              y=df_t3_class_2$Y, 
              z=df_t3_class_2$Z, 
              main = "3D Scatter Plot - Class 2",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_3$X, 
              y=df_t3_class_3$Y, 
              z=df_t3_class_3$Z, 
              main = "3D Scatter Plot - Class 3",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_4$X, 
              y=df_t3_class_4$Y, 
              z=df_t3_class_4$Z, 
              main = "3D Scatter Plot - Class 4",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_5$X, 
              y=df_t3_class_5$Y, 
              z=df_t3_class_5$Z, 
              main = "3D Scatter Plot - Class 5",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_6$X, 
              y=df_t3_class_6$Y, 
              z=df_t3_class_6$Z, 
              main = "3D Scatter Plot - Class 6",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_7$X, 
              y=df_t3_class_7$Y, 
              z=df_t3_class_7$Z, 
              main = "3D Scatter Plot - Class 7",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")

scatterplot3d(x=df_t3_class_8$X, 
              y=df_t3_class_8$Y, 
              z=df_t3_class_8$Z, 
              main = "3D Scatter Plot - Class 8",
              xlab = "X",
              ylab = "Y",
              zlab = "Z")
```

Applying PCA (Subtask b)

```{r}
df_t3_pca <- princomp(df_t3_long %>% select(X,Y,Z), cor = T)
summary(df_t3_pca, loadings = T)
```

Plotting variances of components. From the plot, it can be seen that most of the variance covered by the first 2 components.

```{r}
plot(df_t3_pca, type = "b", main = "Variances of Components")
```

Visualizing each data point on a 2D scatter plot with the new coordinates (different colors for each class). 

Component 1 covers **49.06%** of the variance and Component 2 covers **34.67%** of the variance. Thus, in total, **83.73%** of the variance are covered by the two dimensions.

```{r}
autoplot(df_t3_pca, data = df_t3_long, colour = "class")
```


